Palomar 5 Gaps

### 2024-10-18 paper/main.tex
updated the opening paragraph
resarch rabbit and zotero bib's DONT WORK. Double author. Changing all to ADS. Big pain. Much sad

@@

  The existence of dark matter is supported by substantial observational evidence on scales larger than individual galaxies. This work aims to contribute to the search for dark matter on sub-galactic scales, specifically within the Milky Way. First proposed by [Author Name] in [Year], the concept of dark matter arose to explain the unusually high velocities of stars in dwarf galaxies—velocities that exceed those predicted by the visible mass exerting gravitational forces. Dark matter's influence extends to cosmological scales as well, as demonstrated by the bending of light within cosmic web filaments due to the unseen mass. Despite significant progress, the true nature of dark matter remains elusive, and scientists continue to search for evidence of its existence beyond its gravitational effects on cosmic and extragalactic scales. A key focus today is detecting dark matter on sub-galactic scales, such as through the study of stellar streams.

  Stellar streams, long and thin structures formed by the tidal disruption of globular clusters or dwarf galaxies orbiting a host galaxy, present a promising method for probing dark matter distribution. These tidal forces arise due to differential gravitational pulls across extended objects, causing stars farther from the galactic center to lag behind while those closer are pulled away. This stretching creates two tidal tails that trace the cluster's orbit. Theoretical predictions of this phenomenon existed long before the first observation in 1995 by \citet{carl_j_grillmair_globular_1995}, who detected stars beyond the tidal radii of globular clusters. Subsequent work by \citet{odenkirchen_detection_2001} further demonstrated that Palomar 5's stars not only extended beyond the tidal radius but also formed distinct tidal tails.

  Over the past two decades, theoretical and numerical studies have shown that stellar streams can constrain the gravitational field of host galaxies, particularly the Milky Way. By applying Poisson's equation, the density distribution and gravitational potential can be separated into different components, including the dark matter halo. This halo, often modeled as spherical, is the largest component and crucial to explaining the constant rotational velocity of galaxies. For example, \citet{varghese2011stellar} demonstrated how stellar streams can infer the mass and scale parameters of haloes using various amounts of observational data, from basic right ascension and declination to full six-dimensional phase space information. Meanwhile, Bonaca et al. employed an information-theoretic approach, identifying the orbits and configurations of stellar streams that yield the most information about the galactic potential.

  The Gaia Data Release 2 (DR2) and DR3 revolutionized the observational study of stellar streams, enabling significant advances in the field. Malhan et al. developed the StreamFinder algorithm, which identified 90 stellar streams within the Milky Way, some of which have been followed up by other instruments and surveys. The best-observed streams, such as GD-1 and Palomar 5, contain between 1,000 and 3,000 stars. Leveraging this dataset, Ibata et al. (2024) constructed the most accurate model of the Milky Way to date based on stellar stream data.

  In this study, we build on this body of work by investigating another important use case for stellar streams: probing dark matter on sub-galactic scales. According to simulations by Springel et al. (2008), $\Lambda$-Cold Dark Matter ($\Lambda$CDM) predicts that galaxies grow hierarchically, with dark matter clumps forming at a wide range of masses and sizes. These clumps, or \textit{subhalos}, are predicted to follow a mass distribution with a power-law slope slightly shallower than -2.0. To date, the smallest observed dark matter halo was detected through gravitational lensing in an Einstein ring, as described by Vingetti et al. (20XX). However, some models predict that dark matter clumps could exist down to the mass of a planet like Jupiter (I hear this at conferences, is it true? where are the papers that suggest this?).

  \citet{rodrigo_ibata_uncovering_2002} first suggested that dark matter subhalos could influence stellar streams by diffusing their orbital elements. Later, \citet{r_g_carlberg_pal_2012} expanded this idea, proposing that subhalos might create gaps in stellar streams during flyby encounters, where a subhalo approaches closely enough to a segment of a stream and significantly changes the orbits of the closest stars. This theory was further explored through a probabilistic model of encounter rates based on the expected subhalo distribution. \citet{bonaca_spur_2018} provided observational evidence for this idea, identifying an underdensity in the GD-1 stream that could not be explained by known objects, such as globular clusters, and, interestingly enough, was inconsistent with the $\Lambda$CDM mass-size relationship from M et al 20XX.

>>>>

\newcommand{\paola}[1]{\textcolor{magenta}{{#1}}}
\newcommand{\salvatore}[1]{\textcolor{green}{{#1}}}

@@

  \salvatore{I couldn't find the bib tex for zwicky's original paper}
  Over a century ago, \citet{1933AcHPh...6..110Z} postulated the existence of dark matter to explain the excessively large velocity dispersions of galaxies within their galaxy clusters. Since then, mounting evidence has been assembled to support this claim. \citet{1970IAUS...38..351F,1970ApJ...161..802F} elaborated on the dynamics of disk galaxies and briefly acknoledged in the appendix of his paper that the rotation curves of NGC 300 and M~33 cannot be fully explained by disk models and that there must be the same amount of mass or more that is undetected and beyond the extend of the galaxies to explain their rotation curves. Concurrently \citet{1970ApJ...159..379R} studied the rotation of Andromenda and observed it to be flat and not decreasing as one would expect from a potential with more mass concentration in the center. Eventually, astronomers returned to Zwicky's non-luminous matter to explain these fast-rotating disks, as discussed in \citet{1983SciAm.248f..96R}'s review.

  \salvatore{I see why we cite freeman70 now. there are two comments in the appendix stating that there is as much mass beyond the optical limit as there is within it for two different galaxoes. NGC 300 and M 33. It's not clear to me why we cite Rubin 70. The rotation curve is flat, to me implies a logarithmic potential and hence we need dark matter. However, I don't get any impression of missing matter when I read the text.}

  Beyond the galactic scale, dark matter is essential for explaining many cosmological phenomena. For instance, the Planck mission \citep{2020A&A...641A...1P} was designed to study the Cosmic Microwave Background (CMB), whose spectrum closely follows that of a black body with a peak temperature in the millimeter wavelength range, corresponding to approximately 2.725~K. However, there are slight temperature variations on the order of one part in ten thousand. Notably, the colder regions of the CMB temperature map correlate with whole-sky gravitational lensing maps \citep{2020A&A...641A...8P}. This is because photons lose energy as they escape from gravitational potentials. Additionally, galaxy clusters reside within these deep gravitational wells. The $\Lambda$ Cold Dark Matter model accounts for these observations almost perfectly, proposing that dark matter exists and that its stochastic quantum fluctuations led to density variations. These variations propagated as the universe expanded, shaping the large-scale structure and distribution of galaxies we observe today.

  As laid out in \citet{2021PrPNP.11903865A}'s review, despite significant progress the true nature of dark matter remains elusive and scientists continue to search for evidence of its existence beyond its gravitational effects on cosmic and galactic scales. A key focus today is detecting dark matter on sub-galactic scales, because at these scales different cosmological scenarios predict different lumpiness of the halos. For instance, \citet{2008ApJ...679.1260M} compared the Via Lactea simulations to some that they made themselves and showed that you can have a large variation for the number of dark matter subhalos within a galaxy. \citet{2019Galax...7...81Z} provides a review for the different proposed mechanisms of forming dark matter haloes from perturbation theory and deviations from a uniformly dense medium and also disccses techniques and limitations for modelign such physical phenomena. \citet{2019Galax...7...81Z} rightly states that until particles are discovered $\Lambda$-CDM remains a hypothesis. Since the true nature of dark matter is elusive. The possibilities for sub-galactic behaviour and structure of dark matter are vaste. However, as proposed by Carlberg in 2012, a most promising tool to set limits on the number and mass density of the sub-halos, are stellar streams.

  Stellar streams, several-kpc long structures formed by the tidal disruption of globular clusters or dwarf galaxies orbiting a host galaxy. These tidal forces arise due to differential gravitational pulls across extended objects, causing stars farther from the galactic center to lag behind while those closer are pulled away. This stretching creates two tidal tails that trace the cluster's orbit, unless in the closest vicinity to the object \citep{montuori07}. Numerical predictions of this phenomenon exist since the 1970s \citep[see, for example][]{keenan75}. These predictions occured well before the first detections which were made by \citet{carl_j_grillmair_globular_1995} around some Galactic globular clusters. Interestingly, this work was a nearly contemporary discovery of the Sagittarius stellar stream by \citet{ibata1994} which is the closest example of a stream emerging from a dwarf satellite currently interacting with the Milky Way. Subsequent studies\footnote{\citep[see][for more subsequent observation detections of tidal debris, i.e. globular cluster stars beyond the tidal radius.]{lehmann97, testa00, leon00, siegel01, lee03}} extended Grillmair's findings to other globular cluster streams, until the discovery made by \citet{2001ApJ...548L.165O,odenkirchen02, odenkirchen03} of long and thin tails outside the Palomar~5 globular cluster, which were shown to be extended for more than $10^\circ$ degrees in the sky. These early works stimulated a vigorous search for stellar streams in the Galaxy in the following years \citep{grillmair06a, grillmair06b, belokurov06, grillmair09}. The review of Bonaca properly asserts that we have envolved into a new area of stellar stream astronomy Gaia astrometric mission \citep{prousti16}.

  Gaia's characterization of billions of stars in the Milky Way allows us to look for these structures coupling photometry, astrometry and spectroscopy. The development and deployment of the Streamfinder algorithm has led to the detection of about 90 stellar streams in the Milky Way. \citet{mateu23} provides a summary of these streams and tabulates that about 20 are associated to modern day globular clusters. Thus, currently there are about 70 progenitor-less streams and about 145 stream-less clusteres. The solo-streams can be explained by a larger globular cluster population whose memebers have been disolved due to two body relaxation and tidal stripping. For instance, Z Wan 2020 reports that the Phenoix stream is dynamically consistent from a globular cluster \footnote{see bovy for a description on the difference between dwarf and cluster streams} yet far more metal poor than any cluster ever observed, galactic or otherwise. This implies that the stream is from a completely disolved globular cluster which comes from a population of clusters that have not survived until modern times. Nicolas Martin in 2022 showed that many very metal poor streams exist in the Milky Way. On the other hand, it is possible that some globular clusters do not loose a significant number of stars to create tidal tails. For instance, Vitral and Bouldrini demonstrated that clusters existing in dark matter subhaloes can place the tidal radius far beyond their half mass radii and sheild stars from escape. Another factor is indeed detectability. Additionally, there are over XXX cluster in the internal regional of the galaxy, a region overwhich stream finder was not applied since the high stellar density implicates a computational cost that is currently far to expensive. Also, Balbinot demonstrates that internal dynamics preferentially kicks out small stars, which have lower luminosities and can make detection more difficult.

  One of the interests in studying stellar streams, is that they can constrain the gravitational field of their host galaxies, particularly the Milky Way. For example, \citet{varghese2011stellar} demonstrated how stellar streams can infer the mass and scale parameters of \paola{dark matter} haloes using various amounts of observational data, from basic right ascension and declination to full six-dimensional phase space information. \citet{bonaca14} reviewed this concept from an information-theoretic point of view, identifying which orbits and configurations of stellar streams that yield the most information about the galactic potential. However, the use of single streams for constraining the Galactic potential has led, however, to some ambiguous and not converging results over time. For example, \citet{law10} made use of the Sagittarius stream to infer that the dark matter halo of our Galaxy has a triaxial shape \citep[but see also][]{helmi04, johnston05, law05}; \citet{bovy16} concluded that the dark matter halo of our Galaxy is nearly spherical at the radii of the Palomar~5 and GD~1 streams. Nonetheless, a recent goal of stellar stream astronomy was realized by \citet{ibata24} constructed the most accurate model of the Milky Way to date based on a set of about 90 streams.

  In this study, we build on this body of work by investigating another important use case for stellar streams: probing dark matter on sub-galactic scales. According to simulations by Springel et al. (2008), $\Lambda$-Cold Dark Matter ($\Lambda$CDM) predicts that galaxies grow hierarchically, with dark matter clumps forming at a wide range of masses and sizes. These clumps, or \textit{subhalos}, are predicted to follow a mass distribution with a power-law slope slightly shallower than -2.0. To date, the smallest observed dark matter halo was detected through gravitational lensing in an Einstein ring, as described by \citet{vegetti12}. However, some models predict that dark matter clumps could exist down to the mass of Earth-like planets at least \citep[see][ and discussion in \citet{amorisco21}]{green05, wang20}.

  \citet{rodrigo_ibata_uncovering_2002} first suggested that dark matter subhalos could influence stellar streams by diffusing their orbital elements. Later, \citet{r_g_carlberg_pal_2012} expanded this idea, proposing that subhalos could create gaps in stellar streams during flyby encounters, where a subhalo approaches closely enough to a segment of a stream and significantly changes the orbits of the closest stars. Carlberg 2013 further explored this idea through a probabilistic model of encounter rates based on the expected subhalo distribution. \citet{bonaca_spur_2018} provided observational evidence for this idea, identifying an underdensity in the GD-1 stream that could not be explained by known objects, such as globular clusters, and, interestingly enough, was inconsistent with the $\Lambda$CDM mass-size relationship from M et al 20XX.
